Statistical Guarantees for Local Spectral Clustering on Random Neighborhood Graphs

### 2021-01-27 jmlr/revision/local_spectral_clustering_jmlr_revision_2.tex
Editing pass over the main document complete.

\begin{keywords}
	graphs, spectral clustering, density clustering, Personalized PageRank, unsupervised learning
\end{keywords}

@@

Before giving a more detailed overview of our main results, we review some of the aforementioned worst-case guarantees, formally define PPR on a neighborhood graph, and introduce the population-level functionals that govern the behavior of local clustering in our statistical context.

@@

Once $p_v$ or $p_v^{(\varepsilon)}$ is computed, the cluster estimate $\wh{C}$ is chosen by taking a particular sweep cut of $p_v$. For a given level $\beta > 0$, the \emph{$\beta$-sweep cut} of $p_v = (p_v(u))_{u \in V}$ is 
\begin{equation}
\label{eqn:sweep_cuts}
S_{\beta,v} := \set{u \in V: \frac{p_v(u)}{D_{uu}} > \beta}.
\end{equation}
To determine $\wh{C}$, one computes $S_{\beta,v}$ over all \smash{$\beta \in (L, U)$} (where the range $(L,U)$ is user-specified), and then outputs the cluster estimate
\smash{$\wh{C} = S_{\beta^*,v}$} with minimum normalized cut. For a set $C \subseteq V$ with complement $C^c = V \!\setminus\! C$, the \emph{cut} and \emph{volume} are respectively,
\begin{equation}
\label{eqn:cut_volume}
\cut(C;G) := \sum_{u \in C} \sum_{v \in C^c}
\1\{(u,v) \in E\},~~ \vol(C; G) := \sum_{u \in C}  \sum_{v \in V} \1\{(u,v) \in E\},
\end{equation}
and the \emph{normalized cut} of $C$ is
\begin{equation}
\label{eqn:normalized_cut}
\Phi(C; G) := \frac{\cut(C;G)}{\min \set{\vol(C; G), \vol(C^c; G)}}.
\end{equation}

\subsection{Worst-case guarantees for PPR clustering}
As mentioned, to date most analysis of local clustering has focused on worst-case guarantees, defined with respect to functionals of an a priori fixed graph $G = (V,E)$.  For instance, \cite{andersen2006} analyze the normalized cut of the cluster estimate $\wh{C}$ output by PPR, showing that when PPR is appropriately seeded within a candidate cluster $C \subseteq V$, the normalized cut of $\wh{C}$ is upper bounded $\Phi(\wh{C};G) \lesssim \sqrt{\Phi(C;G)}$.\footnote{We say $a \lesssim b$ when $a \leq $} \cite{zhu2013} build on this: they introduce a second functional, the \emph{conductance} $\Psi(G)$, defined as 
\begin{equation}
\label{eqn:conductance}
\Psi(G) := \min_{S \subseteq V} \Phi(S;G),
\end{equation}
and show that if $\Phi(C;G)$ is much smaller than $\Psi(G[C])^2$---where $G[C] = (C,E \cap (C \times C))$ is the subgraph of $G$ induced by $C$--- then (in addition to having a small normalized cut) the cluster estimate $\wh{C}$ approximately recovers $C$. Our own analysis builds on that of~\cite{zhu2013}, and we give a more detailed summary of their results in Section~\ref{sec:ub_symmetric_set_difference}.
For now, we merely reiterate that the conclusions of \citep{andersen2006,zhu2013} cannot be straightforwardly applied to our statistical setting, where the input data are random samples $\{x_1,\ldots,x_n\}$ drawn from a distribution $\mathbb{P}$, the graph $G$ is a random neighborhood graph formed by the user, and the candidate cluster is a set $\mc{C} \subset \Rd$.\footnote{Throughout, we use calligraphic notation to refer to subsets of $\Rd$.}

\subsection{PPR on a neighborhood graph}
\label{subsec:ppr_neighborhood_graph}
We now formally describe the statistical setting in which we operate, as well as the method we will study: PPR on a neighborhood graph. Let $X = \{x_1,\ldots, x_n\}$ be samples drawn i.i.d.\ from a distribution
$\Pbb$ on $\Rd$. We will assume throughout that $\Pbb$ has a density $f$ with respect to the Lebesgue measure $\nu$ on $\Rd$ . For a radius $r > 0$, we define
$G_{n,r}=(V,E)$ to be the \emph{$r$-neighborhood graph} of $X$, an
unweighted, undirected graph with vertices $V=X$, and an edge $(x_i,x_j) \in
E$ if and only if $\|x_i - x_j\| \leq r$, where $\|\cdot\|$ is the
Euclidean norm. Once the neighborhood graph $G_{n,r}$ is formed, the PPR vector $p_v$ is then computed over $G_{n,r}$, with a resulting cluster estimate $\wh{C} \subseteq X$. The precise PPR algorithm we analyze is summarized in Algorithm~\ref{alg:ppr}.

@@

\subsection{Population normalized cut and conductance}
Next we define two population-level functionals of $\mc{C}$, the normalized cut $\Phi_{\Pbb,r}(\mc{C})$ and conductance $\Psi_{\Pbb,r}(\mc{C})$, which govern the volume of the symmetric set difference $\Delta(\wh{C},\mc{C}[X])$. Let the population-level \emph{cut} of $\mc{C}$ be the expectation (up to a rescaling) of $\cut_{n,r}(\mc{C}[X]) := \cut(\mc{C}[X]; G_{n,r})$,  and likewise let the population-level \emph{volume} of $\mc{C}$ be the expectation (up to a rescaling) of $\vol_{n,r}(\mc{C}[X]) := \vol(\mc{C}[X]; G_{n,r})$; i.e. let
\begin{equation*}
\mathrm{cut}_{\Pbb,r}(\mc{C}) := \int_{\mc{C}} \int_{\mc{C}^c} \1\{\|x - y\| \leq r\} \,d\Pbb(y) \,d\Pbb(x)~~ \mathrm{vol}_{\Pbb,r}(\mc{C}) := \int_{\mc{C}} \int_{\Rd} \1\{\|x - y\| \leq r\} \,d\Pbb(y) \,d\Pbb(x),
\end{equation*}
where $\mc{C}^c := \Rd \!\setminus\! \mc{C}$.
\begin{definition}[Population normalized cut]
	For a set $\mc{C} \subset \Rd$, distribution $\Pbb$ and radius $r > 0$, the \emph{population normalized cut} is
	\begin{equation}
	\label{eqn:population_normalized_cut}
	\Phi_{\Pbb,r}(\mc{C}) := \frac{\mathrm{cut}_{\Pbb,r}(\mc{C})}{\min\{\mathrm{vol}_{\Pbb,r}(\mc{C}), \mathrm{vol}_{\Pbb,r}(\mc{C}^c)\}}.
	\end{equation}
\end{definition}

@@

\begin{definition}[Population conductance]
	For a set $\mc{C} \subset \Rd$, distribution $\Pbb$ and radius $r > 0$, the \emph{population conductance} is
	\begin{equation}
	\label{eqn:population_conductance}
	\Psi_{\mathbb{P},r}(\mc{C}) = \inf_{\mc{S} \subseteq \mc{C}} \Phi_{\wt{\Pbb},r}(\mc{S}).
	\end{equation}
\end{definition}

It is quite natural that $\Phi_{\Pbb,r}(\mc{C})$ and $\Psi_{\Pbb,r}(\mc{C})$ should help quantify the role geometry plays in local spectral clustering. Indeed, by construction these functionals are quite obviously the population-level analogues of the empirical quantities $\Phi_{n,r}(\mc{C}[X]) := \Phi(\mc{C}[X];G_{n,r})$ and $\Psi_{n,r}(\mc{C}[X]) := \Psi(G_{n,r}\bigl[\mc{C}[X]\bigr])$, and as we have already mentioned, these empirical quantities in turn suffice to upper bound the volume of the symmetric set difference. For this reason, similar population level functionals are used by \citep{shi2009,schiebinger2015,garciatrillos19} in the analysis of \emph{global} spectral clustering in a statistical context. We will comment more on the relationship between these works and our own results in Section~\ref{subsec:related_work}. \\

\textcolor{red}{(TODO): Possibly add a section here \emph{giving}---rather than summarizing---our main results.}

\subsection{Summary of results}
A summary of our results (and an outline for this paper) is as follows.
\begin{itemize}
	\item In Section~\ref{sec:ub_symmetric_set_difference}, we show that under certain conditions the sample normalized cut $\Phi_{n,r}(\mc{C}[X])$ and conductance $\Psi_{n,r}(\mc{C}[X])$ are close to their population level counterparts $\Phi_{\Pbb,r}(\mc{C})$ and $\Psi_{\Pbb,r}(\mc{C})$. As a consequence, we find that $\Delta(\wh{C},\mc{C}[X])$ is small whenever $\Phi_{\Pbb,r}(\mc{C})$ is small relative to $\Psi_{\Pbb,r}(\mc{C})^2$, giving the first population-level guarantees for local clustering in a nonparametric statistical context.
\end{itemize}
In Section~\ref{sec:ppr_density_cluster}, we focus on the special case where the candidate cluster $\mc{C} = \mc{C}_{\lambda}$ is a \emph{$\lambda$-density cluster}---that is, a connected component of the upper level set $\{x: f(x) \geq \lambda\}$---an object of central interest in the statistical clustering literature, dating back to the work of \citet{hartigan1981}.
\begin{itemize}
	\item In Section~\ref{subsec:recovery_well-conditioned_density_clusters}, we derive specific upper bounds on the normalized cut $\Phi_{\Pbb,r}(\mc{C}_{\lambda})$ and conductance $\Psi_{\Pbb,r}(\mc{C}_{\lambda})$. These upper bounds depend on $\lambda$ as well as some other natural parameters, and immediately result in an upper bound on $\Delta(\wh{C},\mc{C}[X])$ in the density cluster setting (Theorem~\ref{thm:density_cluster_volume_ssd_ub}).
	\item In Section~\ref{sec:lower_bound}, we prove a negative result: we give a hard distribution $\Pbb$ with corresponding density cluster $\mc{C}_{\lambda}$ for which the symmetric set difference $\Delta(\wh{C},\mc{C}_{\lambda}[X])$ is provably large.
\end{itemize}
In Section~\ref{sec:experiments} we empirically investigate some of our conclusions, before ending with some discussion in Section~\ref{sec:discussion}.

\subsection{Related Work}
\label{subsec:related_work}
We conclude this section by summarizing some related work (in addition to the background already given above).

\paragraph{Spectral clustering.}In the stochastic block model (SBM), arguably one of the simplest models of network formation, edges between nodes independently occur with probability based on a latent community membership. In the SBM, the ability of spectral algorithms to perform clustering---or community detection---is well-understood, dating 
back to \citet{mcsherry2001} who gives conditions under which the entire
community structure can be recovered. In more recent work, \citet{rohe2011}
upper bound the fraction of nodes misclassified by a spectral algorithm for the
high-dimensional (large number of blocks) SBM, and \citet{lei2015} extend these
results to the sparse (low average degree) regime. Relatedly,
\citet{clauset08,balakrishnan2011,li2018}, analyze the misclassification rate
when the block model exhibits some hierarchical structure. The framework we
consider, in which nodes correspond to data points sampled from an underlying 
density, and edges between nodes are formed based on geometric proximity, is
quite different than the SBM, and therefore so is our analysis.

@@

These results are of fundamental importance. However, they remain silent on the following natural question: do the spectra of these continuum operators induce a partition of the sample space which is ``good'' in some sense? \citet{shi2009,schiebinger2015,garciatrillos19} address this question, showing that spectral algorithms will recover the latent labels in certain well-conditioned nonparametric mixture models. These works are probably the most similar to our own: the conditioning of these mixture models depend on population-level functionals resembling the population-level normalized cut and conductance introduced above, and the resulting bounds on the error of spectral clustering are comparable to those we establish in Theorem~\ref{thm:volume_ssd_ub}. However, these results focus on global rather than local methods, and impose global rather than local conditions on $\Pbb$. Moreover, they do not explicitly consider recovery of density clusters, which is an important concern of our work. We comment further on the relationship between our results and these works after Theorem~\ref{thm:volume_ssd_ub}.

\paragraph{Density clustering.} For a given threshold $\lambda \in (0,\infty)$, let \smash{$\mathbb{C}_f(\lambda)$} denote the connected components of the density upper level set $\{x \in \Rd: f(x) > \lambda\}$. In the density clustering problem, initiated by~\cite{hartigan1975}, the goal is to recover $\mathbb{C}_{f}(\lambda)$. By now, density clustering (and the related problem of level-set estimation) are relatively well-understood. \citet{polonik1995,rigollet2009} study density clustering under the symmetric set difference
metric, \citet{tsybakov1997,singh2009} describe minimax optimal level-set
estimators under Hausdorff loss and
\citet{hartigan1981,chaudhuri2010,balakrishnan2013,kpotufe11} consider
consistent estimation of the cluster tree $\{\mathbb{C}_f(\lambda): \lambda \in (0,\infty)\}$. \textcolor{red}{Add, at a minimum, references to Larry + Ale and Ingo's papers.}

We emphasize that our goal is
not to improve on these results, nor to offer a better algorithm for level set
estimation, but rather to better understand the implications of our general theory by applying it within an already well-studied framework. Indeed, seen as a density clustering algorithm, PPR has none 
of the optimality guarantees found in the aforementioned works. We should also note that since we study a local algorithm, our interest will be in a local version of the density clustering problem, where the goal is to recover a single density cluster $\mc{C}_{\lambda} \in \mathbb{C}_f(\lambda)$.

@@

\subsection{Sample-to-population results}
\label{subsec:sample_to_population}
In Propositions~\ref{prop:sample_to_population_1} and~\ref{prop:sample_to_population_2}, we establish high probability bounds on the sample normalized cut, local spread, and conductance, in terms of their population-level analogues. We have already introduced the population normalized cut~\eqref{eqn:population_normalized_cut} and conductance~\eqref{eqn:population_conductance}; denoting $s_{n,r}(\mc{C}[X]) := s\bigl(G_{n,r}\bigl[\mc{C}[X]\bigr]\bigr)$, we now define the population local spread to be
\begin{equation}
\label{eqn:local_spread}
s_{\Pbb,r}(\mc{C}) := \min_{x \in \mc{C}} \biggl\{\frac{\bigl(\deg_{\wt{\Pbb},r}(x)\bigr)^2}{\vol_{\wt{\Pbb},r}(\mc{C})} \biggr\},
\end{equation}
where $\deg_{\wt{\Pbb},r}(x) := \int_{\Rd} \1\{\|y - x\| \leq r\} \,d\wt{\Pbb}(y)$ is the expected degree of $x$ in $G_{n,r}\bigl[\mc{C}[X]\bigr]$.

>>>>

\begin{keywords}
	graphs, spectral clustering, Personalized PageRank, density clustering, unsupervised learning
\end{keywords}

@@

Before giving a more detailed overview of our main results, we formally define PPR on a neighborhood graph, review some of the aforementioned worst-case guarantees, and introduce the population-level functionals that govern the behavior of local clustering in our statistical context.

@@

Once $p_v$ or $p_v^{(\varepsilon)}$ is computed, the cluster estimate $\wh{C}$ is chosen by taking a particular sweep cut. For a given level $\beta > 0$, the \emph{$\beta$-sweep cut} of $p_v = (p_v(u))_{u \in V}$ is 
\begin{equation}
\label{eqn:sweep_cuts}
S_{\beta,v} := \set{u \in V: \frac{p_v(u)}{D_{uu}} > \beta}.
\end{equation}
To determine $\wh{C}$, one computes $S_{\beta,v}$ over all \smash{$\beta \in (L, U)$} (where the range $(L,U)$ is user-specified), and then outputs the cluster estimate
\smash{$\wh{C} = S_{\beta^*,v}$} with minimum normalized cut. For a set $C \subseteq V$ with complement $C^c = V \!\setminus\! C$, the \emph{cut} and \emph{volume} are respectively,
\begin{equation}
\label{eqn:cut_volume}
\cut(C;G) := \sum_{u \in C} \sum_{v \in C^c}
\1\{(u,v) \in E\},~~ \vol(C; G) := \sum_{u \in C}  \sum_{v \in V} \1\{(u,v) \in E\},
\end{equation}
and the \emph{normalized cut} of $C$ is
\begin{equation}
\label{eqn:normalized_cut}
\Phi(C; G) := \frac{\cut(C;G)}{\min \set{\vol(C; G), \vol(C^c; G)}}.
\end{equation}

\subsection{Worst-case guarantees for PPR clustering}
As mentioned, to date most analysis of local clustering has focused on worst-case guarantees, defined with respect to functionals of an a priori fixed graph $G = (V,E)$.  For instance, \cite{andersen2006} analyze the normalized cut of the cluster estimate $\wh{C}$ output by PPR, showing that when PPR is appropriately seeded within a candidate cluster $C \subseteq V$, the normalized cut $\Phi(\wh{C};G)$ is upper bounded by (a constant times)  $\sqrt{\Phi(C;G)}$. \cite{zhu2013} build on this: they introduce a second functional, the \emph{conductance} $\Psi(G)$, defined as 
\begin{equation}
\label{eqn:conductance}
\Psi(G) := \min_{S \subseteq V} \Phi(S;G),
\end{equation}
and show that if $\Phi(C;G)$ is much smaller than $\Psi(G[C])^2$---where $G[C] = (C,E \cap (C \times C))$ is the subgraph of $G$ induced by $C$--- then (in addition to having a small normalized cut) the cluster estimate $\wh{C}$ approximately recovers $C$. Our own analysis builds on that of~\cite{zhu2013}, and we give a more detailed summary of their results in Section~\ref{sec:ub_symmetric_set_difference}.
For now, we merely reiterate that the conclusions of \citep{andersen2006,zhu2013} cannot be straightforwardly applied to our statistical setting, where the input data are random samples $\{x_1,\ldots,x_n\}$ drawn from a distribution $\mathbb{P}$, the graph $G$ is a random neighborhood graph formed by the user, and the candidate cluster is a set $\mc{C} \subset \Rd$.\footnote{Throughout, we use calligraphic notation to refer to subsets of $\Rd$.}

\subsection{PPR on a neighborhood graph}
\label{subsec:ppr_neighborhood_graph}
We now formally describe the statistical setting in which we operate, as well as the method we will study: PPR on a neighborhood graph. Let $X = \{x_1,\ldots, x_n\}$ be samples drawn i.i.d.\ from a distribution
$\Pbb$ on $\Rd$. We will assume throughout that $\Pbb$ has a density $f$ with respect to the Lebesgue measure $\nu$ on $\Rd$ . For a radius $r > 0$, we define
$G_{n,r}=(V,E)$ to be the \emph{$r$-neighborhood graph} of $X$, an
unweighted, undirected graph with vertices $V=X$, and an edge $(x_i,x_j) \in
E$ if and only if $i \neq j$ and $\|x_i - x_j\| \leq r$, where $\|\cdot\|$ is the
Euclidean norm. Once the neighborhood graph $G_{n,r}$ is formed, the PPR vector $p_v$ is then computed over $G_{n,r}$, with a resulting cluster estimate $\wh{C} \subseteq X$. The precise PPR algorithm we analyze is summarized in Algorithm~\ref{alg:ppr}.

@@

\subsection{Population normalized cut and conductance}
Next we define three population-level functionals of $\mc{C}$---the normalized cut $\Phi_{\Pbb,r}(\mc{C})$, conductance $\Psi_{\Pbb,r}(\mc{C})$, and local spread $s_{\Pbb,r}(\mc{C})$---which govern the volume of the symmetric set difference $\Delta(\wh{C},\mc{C}[X])$. Let the population-level \emph{cut} of $\mc{C}$ be the expectation (up to a rescaling) of $\cut_{n,r}(\mc{C}[X]) := \cut(\mc{C}[X]; G_{n,r})$,  and likewise let the population-level \emph{volume} of $\mc{C}$ be the expectation (up to a rescaling) of $\vol_{n,r}(\mc{C}[X]) := \vol(\mc{C}[X]; G_{n,r})$; i.e. let
\begin{equation*}
\mathrm{cut}_{\Pbb,r}(\mc{C}) := \int_{\mc{C}} \int_{\mc{C}^c} \1\{\|x - y\| \leq r\} \,d\Pbb(y) \,d\Pbb(x),~~ \mathrm{vol}_{\Pbb,r}(\mc{C}) := \int_{\mc{C}} \int_{\Rd} \1\{\|x - y\| \leq r\} \,d\Pbb(y) \,d\Pbb(x),
\end{equation*}
where $\mc{C}^c := \Rd \!\setminus\! \mc{C}$. Also let $\deg_{\Pbb,r}(x) := \int_{\Rd} \1\{\|y - x\| \leq r\} \,d\Pbb(y)$ to be the expected degree of $x$ in $G_{n,r}$. 
\begin{definition}[Population normalized cut]
	For a set $\mc{C} \subset \Rd$, distribution $\Pbb$ and radius $r > 0$, the \emph{population normalized cut} is
	\begin{equation}
	\label{eqn:population_normalized_cut}
	\Phi_{\Pbb,r}(\mc{C}) := \frac{\mathrm{cut}_{\Pbb,r}(\mc{C})}{\min\{\mathrm{vol}_{\Pbb,r}(\mc{C}), \mathrm{vol}_{\Pbb,r}(\mc{C}^c)\}}.
	\end{equation}
\end{definition}

@@

\begin{definition}[Population conductance]
	For a set $\mc{C} \subset \Rd$, distribution $\Pbb$ and radius $r > 0$, the \emph{population conductance} is
	\begin{equation}
	\label{eqn:population_conductance}
	\Psi_{\mathbb{P},r}(\mc{C}) = \inf_{\mc{S} \subseteq \mc{C}} \Phi_{\wt{\Pbb},r}(\mc{S}).
	\end{equation}
\end{definition}
\begin{definition}[Population local spread]
	For a set $\mc{C} \subset \Rd$, distribution $\Pbb$ and radius $r > 0$, the \emph{population local spread} is
	\begin{equation}
	\label{eqn:local_spread}
	s_{\Pbb,r}(\mc{C}) := \min_{x \in \mc{C}} \biggl\{\frac{\bigl(\deg_{\wt{\Pbb},r}(x)\bigr)^2}{\vol_{\wt{\Pbb},r}(\mc{C})} \biggr\},
	\end{equation}
\end{definition}

It is quite natural that $\Phi_{\Pbb,r}(\mc{C})$ and $\Psi_{\Pbb,r}(\mc{C})$ should help quantify the role geometry plays in local spectral clustering. Indeed, by construction these functionals are quite obviously the population-level analogues of the empirical quantities $\Phi_{n,r}(\mc{C}[X]) := \Phi(\mc{C}[X];G_{n,r})$ and $\Psi_{n,r}(\mc{C}[X]) := \Psi(G_{n,r}\bigl[\mc{C}[X]\bigr])$, and as we have already mentioned, these empirical quantities in turn suffice to upper bound the volume of the symmetric set difference. For this reason, similar population level functionals are used by \citep{shi2009,schiebinger2015,garciatrillos19} in the analysis of \emph{global} spectral clustering in a statistical context. We will comment more on the relationship between these works and our own results in Section~\ref{subsec:related_work}.

The role played by $s_{\Pbb,r}(\mc{C})$ is somewhat less obvious. For now, we mention only that it plays an essential role in obtaining tight bounds on the mixing time of a particular random walk that can be related to the PPR vector, and defer further discussion until later, in Section~\ref{sec:ub_symmetric_set_difference}.

\subsection{Main Results}
We now informally state our two main upper bounds, regarding the recovery of a generic cluster $\mc{C}$, and a density cluster $\mc{C}_{\lambda}$. Theorem~\ref{thm:volume_ssd_ub_informal} informally summarizes the first of our main results, formally stated in Theorem~\ref{thm:volume_ssd_ub}, regarding the recovery of a generic cluster $\mc{C}$, .
\begin{theorem}[Informal]
	\label{thm:volume_ssd_ub_informal}
	Let $\mc{C} \subset \Rd$ and $\Pbb$ satisfy appropriate regularity conditions, and suppose Algorithm~\ref{alg:ppr} is well-initialized with respect to $\mc{C}$. For all $n \in \mathbb{N}$ sufficiently large, with high probability it holds that
	\begin{equation*}
	\frac{\Delta(\wh{C},\mc{C}[X])}{\vol_{n,r}(\mc{C}[X])} \leq c \cdot\Phi_{\Pbb,r}(\mc{C}) \cdot  \biggl(\frac{\log(1/s_{\Pbb,r}(\mc{C}))}{\Psi_{\Pbb,r}(\mc{C})}\biggr)^2.
	\end{equation*}
\end{theorem}
(In the above, and throughout, $c$ stands for a universal constant that may change from line to line.) Put even more succinctly, we find that $\Delta(\wh{C},\mc{C}[X])$ is small when $\Phi_{\Pbb,r}(\mc{C})$ is small relative to $\Psi_{\Pbb,r}(\mc{C})^2$ (ignoring $\log$ factors). To the best of our knowledge, this gives the first population-level guarantees for local clustering in the nonparametric statistical context.

Theorem~\ref{thm:density_cluster_volume_ssd_ub_informal} informally summarizes the second of our main results, formally stated in Theorem~\ref{thm:density_cluster_volume_ssd_ub}, regarding the recovery of a $\lambda$-density cluster $\mc{C}_{\lambda}$ by PPR. For reasons that we explain later, in Section~\ref{sec:ppr_density_cluster}, our cluster recovery statement will actually be with respect to the $\sigma$-thickened set $\mc{C}_{\lambda,\sigma} := \{x \in \Rd: \mathrm{dist}(x,\mc{C}_{\lambda}) < \sigma\}$, for a given $\sigma > 0$. The upper bound we establish is a function of various parameters that measure the conditioning of both the density cluster $\mc{C}_{\lambda,\sigma}$ and density $f$ for recovery by PPR. We assume that $\mc{C}_{\lambda,\sigma}$ is the image of a convex set $\mc{K}$ of finite diameter $\mathrm{diam}(\mc{K}) \leq \rho < \infty$ under an Lipschitz, Lebesgue-measure-preserving mapping $g$, with Lipschitz constant $L$. We also assume that $f$ is bounded away from $0$ and $\infty$ on $\mc{C}_{\lambda,\sigma}$,
\begin{equation*}
0 < \lambda_{\sigma} \leq f(x) \leq \Lambda_{\sigma} < \infty~~\textrm{for all $x \in \mc{C}_{\lambda,\sigma}$}
\end{equation*}
while satisfying the low-noise condition 
\begin{equation*}
\inf_{y \in \mc{C}_{\lambda,\sigma}} f(y) - f(x) \geq  \theta \cdot \dist(x, \mc{C}_{\lambda,\sigma})^{\gamma}~~\textrm{for all $x$ such that $\dist(x,\mc{C}_{\lambda,\sigma}) \leq r$.}
\end{equation*}
(Here $\dist(x,\mc{C}) := \inf_{y \in \mc{C}} \|y - x\|$.) 
\begin{theorem}[Informal]
	\label{thm:density_cluster_volume_ssd_ub_informal}
	Let $\mc{C}_{\lambda} \subset \Rd$ be a $\lambda$-density cluster of a distribution $\Pbb$ that satisfies appropriate regularity conditions, and suppose Algorithm~\ref{alg:ppr} is well-initialized with respect to $\mc{C}_{\lambda,\sigma}$. For all $n \in \mathbb{N}$ sufficiently large, with high probability it holds that
	\begin{equation}
	\label{eqn:density_cluster_volume_ssd_ub_informal}
	\frac{\Delta(\wh{C},\mc{C}_{\lambda,\sigma}[X])}{\vol_{n,r}(\mc{C}_{\lambda,\sigma})} \leq c \cdot d^4 \cdot \frac{L^2\rho^2}{\sigma r} \cdot \frac{\Lambda_{\sigma}^2 \lambda (\lambda - \theta \frac{r^{\gamma}}{\gamma + 1})}{\lambda_{\sigma}^4} \cdot \log^2\biggl(\frac{\Lambda_{\sigma}^{2/d} L\rho}{\lambda_{\sigma}^{2/d}2r}\biggr).
	\end{equation}
\end{theorem}
Equation~\eqref{eqn:density_cluster_volume_ssd_ub_informal} reveals the separate roles played by geometry and density in the ability of PPR to recover a density cluster. The parameters $L$, $\rho$ and $\sigma$ capture whether $\mc{C}_{\lambda,\sigma}$ is well-conditioned (short and fat) or poorly-conditioned (long and thin) for recovery by PPR. Likewise, the parameters $\lambda_{\sigma}, \Lambda_{\sigma}, \gamma$ and $\theta$ measure whether $f$ is well-conditioned (approximately uniform over the density cluster, and having thin tails outside the density cluster) or poorly conditioned (vice versa). Theorem~\ref{thm:density_cluster_volume_ssd_ub_informal} tells us that if the thickened density cluster $\mc{C}_{\lambda,\sigma}$ is well-conditioned---i.e. $L^2\rho^2/(\sigma r) \approx 1$---and the density $f$ is well-conditioned near $\mc{C}_{\lambda,\sigma}$---i.e. $\Lambda_{\sigma} \approx \lambda \approx \lambda_{\sigma}$ and $(\lambda - \theta r^{\gamma}/(\gamma + 1)) \ll \lambda$---then PPR will approximately recover the (thickened) density cluster.

\subsection{Related Work}
\label{subsec:related_work}
We now summarize some related work (in addition to the background already given above), regarding the theory of spectral clustering, and of density cluster recovery.

\paragraph{Spectral clustering.}
In the stochastic block model (SBM), arguably one of the simplest models of network formation, edges between nodes independently occur with probability based on a latent community membership. In the SBM, the ability of spectral algorithms to perform clustering---or community detection---is well-understood, dating 
back to \citet{mcsherry2001} who gives conditions under which the entire
community structure can be recovered. In more recent work, \citet{rohe2011}
upper bound the fraction of nodes misclassified by a spectral algorithm for the
high-dimensional (large number of blocks) SBM, and \citet{lei2015} extend these
results to the sparse (low average degree) regime. Relatedly,
\citet{clauset08,balakrishnan2011,li2018}, analyze the misclassification rate
when the block model exhibits some hierarchical structure. The framework we
consider, in which nodes correspond to data points sampled from an underlying 
density, and edges between nodes are formed based on geometric proximity, is
quite different than the SBM, and therefore so is our analysis.

@@

These results are of fundamental importance. However, they remain silent on the following natural question: do the spectra of these continuum operators induce a partition of the sample space which is ``good'' in some sense? \citet{shi2009,schiebinger2015,garciatrillos19,hoffmann2019} address this question, showing that spectral algorithms will recover the latent labels in certain well-conditioned nonparametric mixture models. These works are probably the most similar to our own: the conditioning of these mixture models depend on population-level functionals resembling the population-level normalized cut and conductance introduced above, and the resulting bounds on the error of spectral clustering are comparable to those we establish in Theorem~\ref{thm:volume_ssd_ub}. However, these results focus on global rather than local methods, and impose global rather than local conditions on $\Pbb$. Moreover, they do not explicitly consider recovery of density clusters, which is an important concern of our work. We comment further on the relationship between our results and these works after Theorem~\ref{thm:volume_ssd_ub}.

\paragraph{Density clustering.} For a given threshold $\lambda \in (0,\infty)$, let \smash{$\mathbb{C}_f(\lambda)$} denote the connected components of the density upper level set $\{x \in \Rd: f(x) \geq \lambda\}$. In the density clustering problem, initiated by~\cite{hartigan1975}, the goal is to recover $\mathbb{C}_{f}(\lambda)$. By now, density clustering (and the related problem of level-set estimation) are quite well-understood. For instance, \citet{polonik1995,rigollet2009, rinaldo2010, steinwart2015} study density clustering under the symmetric set difference
metric, \citet{tsybakov1997,singh2009,jiang2017} describe minimax optimal level-set and cluster estimators under Hausdorff loss, and
\citet{hartigan1981,chaudhuri2010,kpotufe11,balakrishnan2013,steinwart2017,wang2019} consider
consistent estimation of the cluster tree $\{\mathbb{C}_f(\lambda): \lambda \in (0,\infty)\}$.

We emphasize that our goal is
not to improve on these results, nor to offer a better algorithm for density clustering. Indeed, seen as a density clustering algorithm, PPR has none 
of the optimality guarantees found in the aforementioned works. Rather, we hope to better understand the implications of our general theory by applying it within an already well-studied framework. We should also note that since we study a local algorithm, our interest will be in a local version of the density clustering problem, where the goal is to recover a single density cluster $\mc{C}_{\lambda} \in \mathbb{C}_f(\lambda)$.

\subsection{Organization}
We now outline the rest of the paper.
\begin{itemize}
	\item In Section~\ref{sec:ub_symmetric_set_difference}, we show that under certain conditions the sample normalized cut $\Phi_{n,r}(\mc{C}[X])$ and conductance $\Psi_{n,r}(\mc{C}[X])$ are close to their population level counterparts $\Phi_{\Pbb,r}(\mc{C})$ and $\Psi_{\Pbb,r}(\mc{C})$. As a consequence, we find that $\Delta(\wh{C},\mc{C}[X])$ is small whenever $\Phi_{\Pbb,r}(\mc{C})$ is small relative to $\Psi_{\Pbb,r}(\mc{C})^2$, giving the first population-level guarantees for local clustering in a nonparametric statistical context.
	\item In Section~\ref{sec:ppr_density_cluster}, we focus on the special case where the candidate cluster $\mc{C} = \mc{C}_{\lambda}$ is a \emph{$\lambda$-density cluster}---that is, a connected component of the upper level set $\{x: f(x) \geq \lambda\}$---an object of central interest in the statistical clustering literature, dating back to the work of \citet{hartigan1981}. We derive specific upper bounds on the normalized cut $\Phi_{\Pbb,r}(\mc{C}_{\lambda})$ and conductance $\Psi_{\Pbb,r}(\mc{C}_{\lambda})$. These upper bounds depend on $\lambda$ as well as some other natural parameters, and immediately result in an upper bound on $\Delta(\wh{C},\mc{C}[X])$ in the density cluster setting (Theorem~\ref{thm:density_cluster_volume_ssd_ub}).
	\item In Section~\ref{sec:lower_bound}, we prove a negative result: we give a hard distribution $\Pbb$ with corresponding density cluster $\mc{C}_{\lambda}$ for which the symmetric set difference $\Delta(\wh{C},\mc{C}_{\lambda}[X])$ is provably large.
	\item In Section~\ref{sec:experiments} we empirically investigate some of our conclusions, before ending with some discussion in Section~\ref{sec:discussion}.
\end{itemize}

\textcolor{red}{(TODO): Edit this, based on whether or not the Main Results section passes muster. }

@@

\subsection{Sample-to-population results}
\label{subsec:sample_to_population}
In Propositions~\ref{prop:sample_to_population_1} and~\ref{prop:sample_to_population_2}, we establish high probability bounds on the sample normalized cut, local spread, and conductance, in terms of their population-level analogues. We have already introduced the population normalized cut~\eqref{eqn:population_normalized_cut} and conductance~\eqref{eqn:population_conductance}; denoting $s_{n,r}(\mc{C}[X]) := s\bigl(G_{n,r}\bigl[\mc{C}[X]\bigr]\bigr)$, we now define the population local spread to be
\begin{equation}
\label{eqn:local_spread}
s_{\Pbb,r}(\mc{C}) := \min_{x \in \mc{C}} \biggl\{\frac{\bigl(\deg_{\wt{\Pbb},r}(x)\bigr)^2}{\vol_{\wt{\Pbb},r}(\mc{C})} \biggr\},
\end{equation}
where $\deg_{\wt{\Pbb},r}(x) := \int_{\Rd} \1\{\|y - x\| \leq r\} \,d\wt{\Pbb}(y)$ is the expected degree of $x$ in $G_{n,r}\bigl[\mc{C}[X]\bigr]$ (normalized by a factor of $1/(n - 1)$.)
